something about variable-byte encoding

### 2020-02-08 gradu/gradu.tex
more gradu stuff

\documentclass[english,twoside,censored,tkt,alko-line]{HYthesisML}

@@

%% Reference style: select one 
%% for APA = Harvard style = authoryear -> (Einstein, 1905) use:
\usepackage[style=authoryear,bibstyle=authoryear,backend=biber,natbib=true,maxnames=99,maxcitenames=2,giveninits=true,uniquename=init]{biblatex}
%% for numeric = Vancouver style -> [1] use:
%\usepackage[style=numeric,bibstyle=numeric,backend=biber,natbib=true,maxbibnames=99,giveninits=true,uniquename=init]{biblatex}
%% for alpahbetic -> [Ein05] use:
%\usepackage[style=alphabetic,bibstyle=alphabetic,backend=biber,natbib=true,maxbibnames=99,giveninits=true,uniquename=init]{biblatex}
%

@@

\usepackage{lmodern}         % Font package, again in some systems.
\usepackage{textcomp}        % Package for special symbols
\usepackage[pdftex]{color, graphicx} % For pdf output and jpg/png graphics
\usepackage{epsfig}
\usepackage{subfigure}
\usepackage[pdftex, plainpages=false]{hyperref} % For hyperlinks and pdf metadata
\usepackage{fancyhdr}        % For nicer page headers
\usepackage{tikz}            % For making vector graphics (hard to learn but powerful)
%\usepackage{wrapfig}        % For nice text-wrapping figures (use at own discretion)
\usepackage{amsmath, amssymb} % For better math

\singlespacing               %line spacing options; normally use single

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% STEP 2:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Set up personal information for the title page and the abstract form.
%% Replace parameters with your information.
\title{Tutkielman otsikko}

@@

\begin{abstract}

Tämä dokumentti on tarkoitettu Helsingin yliopiston tietojenkäsittelytieteen osaston opin\-näyt\-teiden ja harjoitustöiden ulkoasun ohjeeksi ja mallipohjaksi. Ohje soveltuu kanditutkielmiin, ohjelmistotuotantoprojekteihin, seminaareihin ja maisterintutkielmiin. Tämän ohjeen lisäksi on seurattava niitä ohjeita, jotka opastavat valitsemaan kuhunkin osioon tieteellisesti kiinnostavaa, syvällisesti pohdittua sisältöä.

Työn aihe luokitellaan  
ACM Computing Classification System (CCS) mukaisesti, 
ks.\ \url{https://www.acm.org/about-acm/class},
käyttäen komentoa \verb+\classification{}+. 
Käytä muutamaa termipolkua (1--3), jotka alkavat juuritermistä ja joissa polun tarkentuvat luokat erotetaan toisistaan oikealle osoittavalla nuolella.

\end{abstract}

\begin{otherlanguage}{english}
\begin{abstract}
Use this otherlanguage environment to write your abstract in another language if needed.

Topics are classified according to the ACM Computing Classification System
(CCS), see 
\url{https://www.acm.org/about-acm/class}:
check command \verb+\classification{}+. A small set of paths (1--3) should be used, starting from any top nodes
referred to bu the root term CCS leading to the leaf nodes. The elements
in the path are separated by right arrow, and emphasis of each element individually can be indicated
by the use of bold face for high importance or italics for intermediate
level. The combination of individual boldface terms may give the reader
additional insight. 
\end{abstract}
\end{otherlanguage}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\backmatter
\begin{appendices}

\input{instructions_english}
%\input{instructions_finnish}

\appendix{Sample Appendix\label{appendix:model}}
usually starts on its own page, with the name and number of the appendix at the top. 
The appendices here are just models of the table of contents and the presentation. Each appendix
Each appendix is paginated separately.

In addition to complementing the main document, each appendix is also its own, independent entity.
This means that an appendix cannot be just an image or a piece of programming, but the appendix must explain its contents and meaning.

\end{appendices}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\documentclass[english,twoside,censored,cs,alko-line]{HYthesisML}

@@

%% Reference style: select one 
%% for APA = Harvard style = authoryear -> (Einstein, 1905) use:
\usepackage[style=authoryear,bibstyle=authoryear,backend=bibtex,natbib=true,maxnames=99,maxcitenames=2,giveninits=true,uniquename=init]{biblatex}
%% for numeric = Vancouver style -> [1] use:
%\usepackage[style=numeric,bibstyle=numeric,backend=biber,natbib=true,maxbibnames=99,giveninits=true,uniquename=init]{biblatex}
%% for alpahbetic -> [Ein05] use:
%\usepackage[style=alphabetic,bibstyle=alphabetic,backend=biber,natbib=true,maxbibnames=99,giveninits=true,uniquename=init]{biblatex}
%

@@

\usepackage{lmodern}         % Font package, again in some systems.
\usepackage{textcomp}        % Package for special symbols
\usepackage[pdftex]{color, graphicx} % For pdf output and jpg/png graphics
\usepackage{epsfig}
\usepackage{subfigure}
\usepackage[pdftex, plainpages=false]{hyperref} % For hyperlinks and pdf metadata
\usepackage{fancyhdr}        % For nicer page headers
\usepackage{tikz}            % For making vector graphics (hard to learn but powerful)
%\usepackage{wrapfig}        % For nice text-wrapping figures (use at own discretion)
\usepackage{amsmath, amssymb} % For better math
\usepackage{algorithm}  
\usepackage[noend]{algpseudocode}
\singlespacing               %line spacing options; normally use single

\algnewcommand\algorithmicforeach{\textbf{for each}}
\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% STEP 2:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Set up personal information for the title page and the abstract form.
%% Replace parameters with your information.
\title{something about variable-byte encoding}

@@

\begin{abstract}
Use this otherlanguage environment to write your abstract in another language if needed.

Topics are classified according to the ACM Computing Classification System
(CCS), see 
\url{https://www.acm.org/about-acm/class}:
check command \verb+\classification{}+. A small set of paths (1--3) should be used, starting from any top nodes
referred to bu the root term CCS leading to the leaf nodes. The elements
in the path are separated by right arrow, and emphasis of each element individually can be indicated
by the use of bold face for high importance or italics for intermediate
level. The combination of individual boldface terms may give the reader
additional insight. 
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\backmatter
%\begin{appendices}

%\input{instructions_english}
%\input{instructions_finnish}

%\end{appendices}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2020-02-08 gradu/msc_english_contents.tex
more gradu stuff

Variable-byte (VB) encoding \citep{Wil99} is a method of compressing integers via omitting leading zero bits. In normal data sets, it loses in 
compression performance to generic algorithms like Huffman encoding or Lempel-Ziv encoding, but in comparison it's faster to decode \citep{Bri09} 
and allows constant time random access. This is later explained in (TODO: link to later chapter).

A good data set for VB encoding is a list of mostly small numbers with a need to support larger ones. A search engine may use an inverted index of 
words in documents. For each word, a list of document IDs where the word appears is stored. It may also store locations of the word in document for 
advanced search purposes. Usually these lists are preprocessed and stored as gaps, storing the difference to previous number instead of the actual 
number \citep{Man08}. Common words have a lot of entries in these lists, but because of gap storing the numbers are small. In contrast, rare words 
have only a few entries but the numbers stored are larger. These lists are excellent data sets for VB encoding.

@@

VB encoding is a well known compression algorithm. It's origins date back to 1980's and the famous MIDI music file format. It stored some of the numbers
in a "variable-length quantity" form, which was a 7-bit block VB structure \citep{Mid96}. Similar data types have existed for example in Apache Lucene 
(as vInt) and IBM DB2 database (as Variable Byte). Later, VB encoding was found efficient in compressing integer lists. It was first used to compress
lists of word locations in documents \citep{Sch02}. It yielded excellent records, and since then many different approaches have been introduced.

history and current implementations etc here
 - mainitse IR book, Lem15, Pla18

TODO: a chapter about existing work

\chapter{Directly addressable codes?}
Rank and select are two functions that work on bit arrays. Rank(i) gives the sum of 1 bits from the beginning of the bit array and select(i) gives
the index of ith 1 bit in the bit array. Both functions work in constant time (citation?) and they require only a few percents of extra space over
the data. The extra bits $c$ added by variable-byte encoding conveniently create a bit array, where 1's represent the endings of numbers. An effective
version of random access has already been introduced \citep{Bri09}.

@@

\chapter{Previous Work}
bl

\chapter{Algorithm}

 - modifications to basic implementation

\chapter{Results}
 - comparison to basic implementation + Bri09
\begin{table}
\centering
\caption{Results with 100k entries (in milliseconds).\label{table:results1}}
\begin{tabular}{l||l c c c c r} 
Experiment & 128 & 256 & 32768 & 65536 & $2^{24}$ & $2^{32} -1$\\ 
\hline \hline 
$7bit VByte encoding$ & 34.97 & 49 & 53.04 & 52.18 & 53.08 & 76.21\\
$8bit VByte encoding$ & 33.57 & 32.47 & 42.96 & 43.11 & 46.15 & 65.14\\
$7bit VByte encoding with array$ & 33.39 & 46.85 & 51.24 & 49.03 & 48.93 & 66.84 \\
$8bit VByte encoding with array$ & 32.52 & 31.88 & 41.54 & 39.94 & 41.15 & 52.86 \\

>>>>

Variable-byte (VB) encoding \citep{Wil99} is a method of compressing integers via omitting leading zero bits. In normal data sets, it loses in 
compression performance to generic algorithms like Huffman encoding or Lempel-Ziv encoding, but in comparison it's faster to decode \citep{Bri09} 
and sometimes preferred due to it's simplicity. allows constant time random access. This is later explained in (TODO: link to later chapter).

A good data set for VB encoding is a list of mostly small numbers with a need to support larger ones. A search engine may use an inverted index of 
words in documents. For each word, a list of document IDs where the word appears is stored. It may also store locations of the word in document for 
advanced search purposes. Usually these lists are preprocessed and stored as an inverted index or gaps, storing the difference to previous number instead of the actual 
number \citep{Man08}. Common words have a lot of entries in these lists, but because of gap storing the numbers are small. In contrast, rare words 
have only a few entries but the numbers stored are larger. These lists are excellent data sets for VB encoding.

@@

VB encoding is a well known compression algorithm. It's origins date back to 1980's and the famous MIDI music file format. It stored some of the numbers
in a "variable-length quantity" form, which was a 7-bit block VB structure \citep{Mid96}. Similar data types have existed for example in Apache Lucene 
(as vInt) and IBM DB2 database (as Variable Byte). Later, VB encoding was found efficient in compressing integer lists. It was first experimented in 
compressing inverted index lists of word locations in documents \citep{Sch02}. That yielded excellent records, and since then many different approaches 
have been introduced.

Modern studies have looked into machine code for VB and applied SIMD (Single instruction, multiple data) architecture to VB \citep{Lem18,Pla15}. The bit 
operations in VB are simple and therefore modifying the code to use SIMD instructions is simple and the speed improvements are significant.

TODO: might need 1-2 more chapters

\chapter{Directly addressable codes?}

Ability to handle large amounts of data fast is one of the key challenges in the field of search engines. Compressed data structures are applied to fit the 
data into cache, memory or even hard drive. Direct access to any element is one of the usual requirements in compressed data strutures. It is not possible
to decode i-th element in known well-performing compression algorithms like Huffman or Lempel-Ziv, because the position of the element depends on the data 
before it. While VB encoding doesn't natively support direct access, it is achievable with two succinct data structure operations: rank and select.

<explain rank>

<explain select>

<introduce Bri09>

<example of Bri09>

- TODO: find out where rank+select initially come from.

Rank and select are two functions that work on bit arrays. Rank(i) gives the sum of 1 bits from the beginning of the bit array and select(i) gives
the index of ith 1 bit in the bit array. Both functions work in constant time (citation?) and they require only a few percents of extra space over
the data. The extra bits $c$ added by variable-byte encoding conveniently create a bit array, where 1's represent the endings of numbers. An effective
version of random access has already been introduced \citep{Bri09}.

@@

\chapter{Rearranged VB codes}
bl

\chapter{Experimental results}
 - comparison to basic implementation + Bri09
\begin{table}
\centering
\caption{Results with 100k entries (in milliseconds).\label{table:results1}}
\begin{tabular}{l||l c c c c r} 
Experiment & 128 & 256 & 32768 & 65536 & $2^{24}$ & $2^{32} -1$\\ 
\hline \hline 
$7bit VByte encoding$ & 34.97 & 49 & 53.04 & 52.18 & 53.08 & 76.21\\
$8bit VByte encoding$ & 33.57 & 32.47 & 42.96 & 43.11 & 46.15 & 65.14\\
$7bit VByte encoding with array$ & 33.39 & 46.85 & 51.24 & 49.03 & 48.93 & 66.84 \\
$8bit VByte encoding with array$ & 32.52 & 31.88 & 41.54 & 39.94 & 41.15 & 52.86 \\
